---
title: "Irish election 2016 and 2020 study"
output:
  pdf_document:
    df_print: kable
    number_sections: yes
    toc: yes
  word_document:
    toc: yes
header-includes: \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
```

\pagebreak
# Introduction
This report presents a study of the Irish election 2020 for the region Galway West and it's comparison with 2016 elections for the same region. It summarizes:  
- The vote per party in each election.  
- The change in votes per party from 2016-2020  
- A comparison to the national average for party share of the vote  
- The change in vote for the most significant candidates in both elections 

```{r}
#Load all the relevant libraries
library(knitr)
library(kableExtra)
library(ggplot2)
library(tidyverse)
library(scales)
library(ggthemes)
library(stringr)
library(gridExtra)
library(colorblindr)



#Load data for 2016 elections
data2016 <-
  read.csv(
    "./GW-election2016-2020/2016-04-28_general-election-count-details-galway-west-csv_en.csv",
    fileEncoding = "UTF-8-BOM"
  )
#Load data for 2020 elections
data2020 <-
  read.csv("./GW-election2016-2020/2020Election.csv", fileEncoding = "UTF-8-BOM")
```


# Votes per party  

Figure 1 gives us insights about the number of votes per party in the 2016 and 2020 Irish General election for Galway-west constituency.

```{r}
#DATA CLEANING -2016 data
#Join the first and surname to get the Candidate name
data2016$Candidate <-
  paste(data2016$Candidate.First.Name,
        " ",
        data2016$Candidate.surname)
#Remove any whtespaces in the party and candidate names
data2016$Party <- as.factor(str_squish(data2016$Party))
data2016$Candidate <- as.factor(str_squish(data2016$Candidate))
#Get the summation of votes per candidate
cleaned_2016_data <-
  data2016 %>% group_by(Party, Candidate) %>% summarize(total_votes = first(Votes))
#Get the summation of votes per party
data_2016_per_party <-
  cleaned_2016_data %>% group_by(Party) %>% summarize(votes_per_party =
                                                        sum(total_votes))
```


```{r}
#DATA CLEANING -2020 data
#Remove any "," in the First preferential vote column
data2020$total_votes <- as.numeric(gsub(",", "", data2020$Count.1))
#Remove any whtespaces in the party and candidate names
data2020$Party <- as.factor(str_squish(data2020$Party))
data2020$Candidate <- as.factor(str_squish(data2020$Candidate))
#Drop the Count.1 column
cleaned_2020_data <- data2020[-3]
#Get the summation of votes per party
data_2020_per_party <-
  cleaned_2020_data %>% group_by(Party) %>% summarize(votes_per_party = sum(total_votes))
```

```{r}
#Change labels of those parties that have been named differently in 2016 data 
levels(data_2016_per_party$Party)[levels(data_2016_per_party$Party) == "Ind"] <-
  "Independent"
#Remove encoding of party labels in 2020 data
levels(data_2020_per_party$Party)[levels(data_2020_per_party$Party) == "Fianna Fáil"] <-
  "Fianna Fail"
levels(data_2020_per_party$Party)[levels(data_2020_per_party$Party) == "Sinn Féin"] <-
  "Sinn Fein"


#Overall use of politically sensitive colour scheme
party.colours <-
  c(
    'Fine Gael' = '#1f78b4',
    'Fianna Fail' = '#33a02c',
    'Sinn Fein' = 'darkolivegreen',
    'Labour Party' = '#e31a1c',
    'Green Party' = '#b2df8a',
    'Social Democrats' = '#cab2d6',
    'Independent' = 'darkred',
    'Renua' = '#ff7f00',
    'Direct Democracy Ireland' = 'darkgrey',
    "Solidarity–PBP" = '#fdbf6f',
    "Aontu" = '#fb9a99',
    "Social Democratic Party" = '#cab2d6',
    "AAA" = '#fdbf6f'
  )
```


```{r, fig.cap="Number of votes per party"}
#Function that returns a bar chart based on the party colour that it belongs to
plot_graph <- function(data, title) {
  #Set the theme as classic
  theme_set(theme_classic())
  #Plot a graph that takes in data and title as parameters and returns a bar chart
  ggplot(data, (aes(
    x = reorder(Party, votes_per_party) ,
    y = votes_per_party,
    fill = reorder(Party, votes_per_party)
  ))) + geom_col(width = 0.8) +
    scale_fill_manual(values = party.colours) +
    ggtitle(title) +
    coord_flip(clip = "off") +
    theme(
      axis.title = element_blank(),
      axis.line.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.line.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_blank(),
      axis.title.x.bottom =  element_blank(),
      legend.position = "none",
      plot.title = element_text(size = 11),
      plot.margin = margin(3, 6, 3, 3),
      panel.background = element_blank(),
      panel.grid.major.x = element_line(
        size = 0.4,
        linetype = 'solid',
        colour = "white"
      ),
      panel.ontop = TRUE
    )
}
#Plot 1 of data 2016 for votes per party
plot1 <- plot_graph(data_2016_per_party, "Votes per party in 2016")
#Plot 2 of data 2020 for votes per party
plot2 <- plot_graph(data_2020_per_party, "Votes per party in 2020")
#Arrange the 2 plots in one line
grid.arrange(plot1, plot2, nrow = 1)
```


## Design decisions  

- The labels of parties are different in 2016 in comparison to the 2020 dataset. These have been modified to be the same as that in the 2020 dataset. Example : Ind -> Independent
- As the chart is color-coded, it is easy to compare the relative performance of parties in the elections. Those parties that have been renamed in 2020, have the same color codes which makes it easier to identify.
- The grid lines are colored white to make it easier to read the horizontal axis which is indicative of the count of votes (FPV)
- Co-ordinates are flipped to make comparison of parties easier and also to enhance clarity

# The change in votes per party from 2016-2020  

The FPV votes for the various parties between 2016 & 2020 elections are not the same. There is a need to make party-wise comparisons for the consecutive elections, this comparison is shown as a divergent bar graph in Figure 2.
The **difference in the FPV gained by each party between the consecutive elections** is calculated and it is this value that is plotted in Figure 2


```{r}
#DATA CLEANING
#Rename parties that have changed their names for 2020 general elections
levels(data_2016_per_party$Party)[levels(data_2016_per_party$Party) == "Social Democratic Party"] <-
  "Social Democrats"
levels(data_2016_per_party$Party)[levels(data_2016_per_party$Party) == "AAA"] <-
  "Solidarity–PBP"

#Add year to each dataset - to differentiate the 2 datasets
data_2016_per_party$year <- as.factor(2016)
data_2020_per_party$year <- as.factor(2020)

#Append the datasets to form a master dataset of (Party, votes_per_party, year)
votes_per_party <- rbind(data_2016_per_party, data_2020_per_party)
#To find the difference, make the data untidy
untidy_data <-
  votes_per_party %>% spread(key = year,
                             value = votes_per_party,
                             convert = FALSE)
#Rename the columns to avoid .X and .Y
colnames(untidy_data) <- c("Party", "Year2016", "Year2020")
#If a value is not defined, set to 0 so that numerical subtraction results are correct
untidy_data[is.na(untidy_data)] <- 0
#Subtract the votes for the years 2020 and 2016 and subset only Party and diff_in_votes columns
diff_per_party <-
  untidy_data %>% mutate(diff_in_votes = Year2020 - Year2016) %>% select("Party", "diff_in_votes")

```

 
```{r,fig.cap="Change in vote per party from 2016 to 2020"}
#Produce a plot which uses party as x-axis, diff_in_votes as y and fills by party for easily comparison
ggplot(diff_per_party, aes(x = Party, y = diff_in_votes, fill = Party)) +
  geom_col(position = "identity",
           alpha = 0.85,
           width = 0.7) +
  scale_fill_manual(values = party.colours) +
  scale_y_continuous(
    breaks = seq(-15e3, 16e3, by = 2e3),
    expand = c(0, 0),
    labels = seq(-15e3, 16e3, by = 2e3),
    name = "difference in votes per party"
  ) +
  theme(
    axis.line.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.x = element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 60, hjust = 1),
    plot.margin = margin(6, 6, 3, 3),
    panel.background = element_blank(),
    panel.grid.major.y = element_line(
      size = 0.2,
      linetype = 'solid',
      colour = "white"
    ),
    panel.ontop = TRUE,
    legend.position = "none"
    
  ) +
  #Add a horizontal line at the origin to differentiate between positive and negative values
  geom_hline(yintercept = 0,
             size = 0.1,
             colour = "grey")
```

## Design Decisions

- The party labels are tilted to an angle of 60 deg to make them readable
- The bar chart uses a politically sensitive color coding scheme
- The ticks on the x-axis help the reader to connect the party name with the correct bar

# A comparison to the national average for party share of the vote

For the 2020 Irish general election, data was collected from www.irelandelection.com This was then loaded as a CSV to calculate the percentage of FPV secured by each party.
This CSV contains 2 columns: Party name (Party) and the total number of FPV per party(total_votes)

```{r}
#Read the CSV of national averages for 2020
data_average_2020 <-
  read.csv("./GW-election2016-2020/National_Avg_2020.csv",
           fileEncoding = "UTF-8-BOM")
#Set the type to Ireland (as it is national average)
data_average_2020$type <- "Ireland"
#Calculate percentage of votes for each of the party for national average dataset
data_average_2020$percentage <-
  round(data_average_2020$total_votes * 100 / sum(data_average_2020$total_votes),
        0)
```

```{r}
#Calculate percentage of votes for each of the party for 2020 datset
data_2020_per_party$percentage <-
  round(
    data_2020_per_party$votes_per_party * 100 / sum(data_2020_per_party$votes_per_party),
    0
  )
#SEt the type to the Galway-west constituency
data_2020_per_party$type <- "Galway-West"
#Append both the datasets to form a master dataset of Party, type and Percentage
national_avg_comp_2020 <-
  rbind(data_average_2020[, c(1, 3, 4)], data_2020_per_party[, c(1, 4, 5)])
```


```{r, fig.cap="Comparison of national average to the constituency for party votes in 2020"}
#CVD plot for 2020 national average comparison to the Galway-West votes per party
ggplot(national_avg_comp_2020 , mapping = aes(x = percentage, y = Party)) +
  geom_line(aes(group = Party), colour = "grey", size = 0.5) +
  geom_point(aes(colour = type), size = 3, alpha = 0.7) +
  #Set the x-axis scale
  scale_x_continuous(
    limits = c(0, 32),
    expand = c(0, 0),
    breaks = seq(0, 32, by = 2),
    name = "Percentage  of FP votes "
  ) +
  #Set the theme
  theme(
    axis.title.y = element_blank(),
    panel.grid.major.x = element_line(size = 0.03),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.line.y = element_blank(),
    axis.line.x = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = c(0.85, 0.94),
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.7, "lines"),
    legend.title = element_blank()
  ) +
  #Add Okabi Ito colour palette
  scale_color_OkabeIto(order = c(1, 5))
```

## Design decisions

- The percentage of the FPV votes secured by each party is plotted on the horizontal axis  
- The political party names are present on the vertical axis, this enhances the readability  
- The points are color-coded in CVD using the **Okabi Ito** color palette to represent the Galway-west percentage and the national average percentage distinctly  
- The 2 unique points for each of the parties are connected by a grey line to help illustrate the magnitude of their difference  
- The major grid lines are used to help get the exact value of the share of the FPV for each point  
- This plot not only showcases the difference in the votes between constituency and national average but also uses less ink. Hence, it is preferable to use dot plots for comparisons.

# The change in votes for the most significant candidates in both elections

In this report, the most significant candidates are those who contested both of the elections.
Mairead Farell was a popular candidate who was elected with 8464 votes which was the second highest in the 2020 election. Hence she has also been included as a significant candidate.  

Figure 5 displays the FPV vote count of the most significant candidates in both elections under comparison

```{r}
#CLEANING DATA
#Some candidates have changes in their names and hence their names are converted to have the same spelling
levels(cleaned_2016_data$Candidate)[levels(cleaned_2016_data$Candidate) ==
                                      "Éamon O'Cuív"] <- "Éamon Ó Cuív"
levels(cleaned_2016_data$Candidate)[levels(cleaned_2016_data$Candidate) ==
                                      "Sean Kyne"] <- "Seán Kyne"
levels(cleaned_2016_data$Candidate)[levels(cleaned_2016_data$Candidate) ==
                                      "Catherine Martina Ann Connolly"] <- "Catherine Connolly"
levels(cleaned_2016_data$Candidate)[levels(cleaned_2016_data$Candidate) ==
                                      "Niall O' Tuathail"] <- "Niall Ó Tuathail"

#Order the data by number of votes (FPV)
cleaned_2016_data <-
  cleaned_2016_data[order(-cleaned_2016_data$total_votes), ]
cleaned_2020_data <-
  cleaned_2020_data[order(-cleaned_2020_data$total_votes), ]

#Get the candidates present in both elections
candidates_compares <-
  merge(cleaned_2016_data, cleaned_2020_data, by = "Candidate")$Candidate
#Subset the significant candidates in 2016 data
candidates_2016 <-
  subset(
    cleaned_2016_data,
    Candidate %in% candidates_compares,
    select = c(Candidate, total_votes)
  )
#Add Mairead as an extra significant candidate to the candidate list
#Set the FPV of Mairead to 0 as she did not contest in the 2016 elections
candidates_2016 <-
  rbind(candidates_2016, as.data.frame(list(
    Candidate = as.factor("Mairéad Farrell"),
    total_votes = 0
  )))
#Set the year as 2016
candidates_2016$year <- as.factor(2016)

#Subset the significant candidates in 2016 data
candidates_2020 <-
  subset(
    cleaned_2020_data,
    Candidate %in% candidates_compares,
    select = c(Candidate, total_votes)
  )
#Add Mairead as an extra significant candidate to the candidate list
candidates_2020 <-
  rbind(candidates_2020, cleaned_2020_data[cleaned_2020_data$Candidate ==
                                             "Mairéad Farrell", c("Candidate", "total_votes")])
#Set the year as 2020
candidates_2020$year <- as.factor(2020)

#Append the 2 candidates datasets from 2016 and 2020
final_candidates_list <- rbind(candidates_2016, candidates_2020)
```


```{r, fig.cap="Change in vote for most significant candidates from 2016 to 2020"}
#Plot of the significant candidates change in votes form 2016 to 2020
ggplot(final_candidates_list , mapping = aes(x = total_votes, y = Candidate)) +
  geom_line(aes(group = Candidate), colour = "grey", size = 0.5) +
  geom_point(aes(colour = year), size = 3, alpha = 0.7) +
  xlab("Number of FP votes") +
  #Set theme
  theme(
    axis.title.y = element_blank(),
    panel.grid.major.x = element_line(size = 0.03),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.line.y = element_blank(),
    axis.line.x = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = c(0.84, 0.84),
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.7, "lines"),
    legend.title = element_blank()
  ) +
  #Viridis colour palette
  scale_color_viridis_d()
```

## Design Decisions

- The number of the FP(First Preferential) votes secured by each candidate is plotted on the horizontal axis  
- The  names of the candidate are present on the vertical axis to enhance the readability  
- The points are color-coded to represent the 2016 & 2020 general elections separately using the **Viridis** color palette.
- The 2 unique points for each of the candidates are connected by a grey line to help point out the difference in votes that each candidate had in comparison between the years 2016 and 2020   
- Mairead Farell, did not contest the 2016 election and hence her FPV in 2016 is set to 0


# Conclusion
- The major parties in both elections in Galway-West constituency can be identified as Fine Gael, Fine Faile, Sinn Fein    
- Fine Fael and Fine Gael show a significant decrease in the number of votes they obtained in 2020 elections in comparison with 2016 elections  
- Sinn Fein and Green Party show a significant increase in the number of votes they obtained in 2020 elections in comparison with 2016 elections  
- Points that are close together for a party in Figure 3, show equivalent proportions of national average and constituency party share of votes  
- Sinn Fein has a larger contribution nationally in comparison to Galway-West  
- Independent parties have a larger contribution in Galway-West in comparison to Ireland  
- Mairead has the largest change in vote as she did not contest in 2016 general elections  
- Most of the candidates improved on the number of votes recieved in 2020 elections except for Sean Kyne and Eamon O Cuiv  

# References
- Lecture slides
- GGPLOT2 Notes
- https://www.rte.ie/news/election-2016/
- https://www.rte.ie/news/election-2020/
- https://en.wikipedia.org/wiki/2020_Irish_general_election
- https://en.wikipedia.org/wiki/2016_Irish_general_election  


# Appendix
```{r eval=FALSE, echo=TRUE, ref.label=knitr::all_labels()}

```



```{r}
total_seats <- sum(data_2016_per_party$votes_per_party)
ggplot(data_2016_per_party, aes(x = "2020", y = votes_per_party, fill = Party))+
  geom_col(width = 0.1) +
  geom_text(aes(label = Party),
            position = position_stack(vjust = 0.5), size=2.6, colour = "white") +
  
  scale_y_continuous(breaks = seq(0,160, by = 20), expand=c(0,0), sec.axis = sec_axis(~ ./ total_seats, labels = scales::percent,
                                                                                      name = "Percentage"))+
  scale_x_discrete( expand = c(0, 0))+
  
  
  #theme_economist(base_size = 8) +
  scale_fill_manual(values = party.colours)+
  theme(legend.position = "None", 
        legend.title = element_blank()) +
  theme(axis.title.y = element_text(margin = margin(r = 10)),
        axis.title.x = element_blank(),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.text.x =  element_blank(),
        axis.ticks.x =  element_blank()) +
  #theme(aspect.ratio = .2) +
  ylab("Seats") +
  xlab(NULL)

```
